Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-01-25 Master Thesis/synopsis.tex
Simplify 5.2

Let $n\in\mathbb{N}$. Tensoring the exact sequence $0\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
\[ 1\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
We'll use the same notation for $P' = \sigma(P)\in X'_0$.

@@

\begin{example}
\begin{itemize}
\item Free or finite abelian groups are free-by-finite.
\item Finitely generated abelian groups are free-by-finite by the structure theorem for finitely generated $\mathbb{Z}$-modules.
\end{itemize}
\end{example}

\begin{lemma}\phantomsection\label{5.2}
\begin{enumerate}[(i)]
\item Let $f: A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite iff $B$ is free-by-finite.
\item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups. If $A'$ and $A''$ are free-by-finite, then $A$ is free-by-finite. 
\item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups. If $A$ is free-by-finite, then $A'$ is free-by-finite. Additionally, if $A'$ is finitely generated, then $A''$ is free-by-finite.
\item Let $A$ be free-by-finite. Then $\bigcap_{n\in\mathbb{N}}nA=0$.
\end{enumerate}
\end{lemma}

@@

Observe the following commutative diagram with exact rows, where we consider $A'$ as a subgroup of $A$:
\[ \begin{tikzcd}
0 \ar[r]& A'/A'_\text{tor}\ar[r]\ar[d, "f"'] & A/A_\text{tor}\ar[r]\ar[d, "\cong"] & A/(A'+A_\text{tor}) \ar[r]\ar[d, "g"]& 0\\
0 \ar[r] & p^{-1}(A''_\text{tor})/A_\text{tor} \ar[r]& A/A_\text{tor}\ar[r] & A''/A''_\text{tor}\ar[r] & 0
\end{tikzcd} \]
Note that $f$ is injective and $g$ is surjective. Using the snake lemma, we get an isomorphism $\operatorname{coker}(f)\cong \ker(g)$. We have:
\[\ker(g) = \ker( (A/A')/(A_\text{tor}/A'_\text{tor}) \to A''/A''_\text{tor} ) = A''_\text{tor}/(A_\text{tor}/A'_\text{tor})\]
Since $A''_\text{tor}$ is finite, $\operatorname{coker}(f)\cong \ker(g)$ is finite, too. Applying (i) on $f$, we can see that $p^{-1}(A''_\text{tor})/A_\text{tor}$ is free-by-finite. Since $A''/A''_\text{tor}$ is free abelian, the short exact sequence in the second row splits, making $A/A_\text{tor}$ a direct sum of a free-by-finite group and a free group.
\item Subgroups of free-by-finite groups are free-by-finite. Now assume $A'$ is finitely generated. We consider $A'$ as a subgroup of $A$. Let $T=A_\text{tor}$. By replacing $A'$, $A$, $A''$ with $A'/(T\cap A')$, $A/T$, $A''/p(T)$ respectively, we can assume that $A$ is free and $A'$ is free of finite rank. Let $B$ be a basis of $A$. There exists a finite subset $C\subset B$ such that $A'\subset \bigoplus_{b\in C}\mathbb{Z}b$. We get an exact sequence:
\[ 0 \longrightarrow \Big(\bigoplus_{b\in C} \mathbb{Z}b\Big)\Big/A' \longrightarrow A/A' \longrightarrow \bigoplus_{b\in B\setminus C}\mathbb{Z}b \longrightarrow 0 \]
The left abelian group is finitely generated, particularly free-by-finite. The right group is free, hence also free-by-finite. By (ii), $A/A'\cong A''$ is free-by-finite as well.
\item It suffices to prove this assertion for finite groups and for $A=\mathbb{Z}$. For $A$ finite we have $nA = 0$ for $n = \#A$. Since $\mathbb{Z}$ has no nilpotent elements, the intersection of all prime ideals of $\mathbb{Z}$ is already trivial. \qedhere
\end{enumerate}
\end{proof}

@@

\begin{proof}
Let $U = X\setminus S$ and $P_1\in U$ be a closed point. We set:
\[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
Since $r_{P_1}|_{k^\times}:k^\times \to k(P_1)^\times$ is injective, we have $C_1\cap k^\times = 1$ and the following map is injective:
\[ C_1 \to\prod_{P\in S}\mathbb{Z},\ x \mapsto (v_P(x))_{P\in S} \]
Therefore, $C_1$ is a free abelian group of finite rank. If $C_1 = 1$, then we set $T = \{P_1\}$ and we are done. Otherwise, let $x\in C_1,\ x\neq 1$. Then there exists an $a\in k^\times$ with the following properties:
\begin{enumerate}[(i)]
\item There exists a point $P_2\in X_0\setminus(S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
\item $a$ is not a root of unity.
\end{enumerate}
Clearly, (i) and (ii) hold for almost all $a\in k^\times$. We set $C_2 = \ker(r_{P_2}|_{C_1})\subset C_1$. $C_2$ is free abelian of finite rank as well. By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$. By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
\[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism:
\[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]
Since $K^S/k^\times$ is finitely generated, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
\end{proof}

>>>>

Let $n\in\mathbb{N}$. Tensoring the exact sequence $1\to U_P\to K^\times \to\mathbb{Z}\to 0$ with $\mathbb{Z}/n\mathbb{Z}$ preserves its exactness since $\mathbb{Z}$ is $\mathbb{Z}$-flat:
\[ 1\longrightarrow U_P/n\longrightarrow K^\times/n\longrightarrow\mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
We'll use the same notation for $P' = \sigma(P)\in X'_0$.

@@

\begin{example}
\begin{itemize}
\item Free or finite abelian groups are free-by-finite.
\item Finitely generated abelian groups are free-by-finite by the structure theorem for finitely generated $\mathbb{Z}$-modules.
\item Any subgroup of a free-by-finite group is free-by-finite as well.
\end{itemize}
\end{example}

\begin{lemma}\phantomsection\label{5.2}
\begin{enumerate}[(i)]
\item Let $f: A\to B$ be a homomorphism of abelian groups with finite kernel and cokernel. Then $A$ is free-by-finite iff $B$ is free-by-finite.
\item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups. If $A'$ and $A''$ are free-by-finite, then $A$ is free-by-finite. 
\item Let $0\longrightarrow A'\stackrel{i}{\longrightarrow}A\stackrel{p}{\longrightarrow}A''\longrightarrow 0$ be an exact sequence of abelian groups. If $A$ is free-by-finite and $A'$ is finitely generated, then $A''$ is free-by-finite.
\item Let $A$ be free-by-finite. Then $\bigcap_{n\in\mathbb{N}}nA=0$.
\end{enumerate}
\end{lemma}

@@

Observe the following commutative diagram with exact rows, where we consider $A'$ as a subgroup of $A$:
\[ \begin{tikzcd}
0 \ar[r]& A'/A'_\text{tor}\ar[r]\ar[d, "f"'] & A/A_\text{tor}\ar[r]\ar[d, "\cong"] & A/(A'+A_\text{tor}) \ar[r]\ar[d, "g"]& 0\\
0 \ar[r] & p^{-1}(A''_\text{tor})/A_\text{tor} \ar[r]& A/A_\text{tor}\ar[r] & A''/A''_\text{tor}\ar[r] & 0
\end{tikzcd} \]
Note that $f$ is injective and $g$ is surjective. Using the snake lemma, we get an isomorphism $\operatorname{coker}(f)\cong \ker(g)$. We have:
\[\ker(g) = \ker( (A/A')/(A_\text{tor}/A'_\text{tor}) \to A''/A''_\text{tor} ) = A''_\text{tor}/(A_\text{tor}/A'_\text{tor})\]
Since $A''_\text{tor}$ is finite, $\operatorname{coker}(f)\cong \ker(g)$ is finite, too. Applying (i) on $f$, we can see that $p^{-1}(A''_\text{tor})/A_\text{tor}$ is free-by-finite. Since $A''/A''_\text{tor}$ is free abelian, the short exact sequence in the second row splits, making $A/A_\text{tor}$ a direct sum of a free-by-finite group and a free group.
\item We consider $A'$ as a subgroup of $A$. Let $T=A_\text{tor}$. By replacing $A'$, $A$, $A''$ with $A'/(T\cap A')$, $A/T$, $A''/p(T)$ respectively, we can assume that $A$ is free and $A'$ is free of finite rank. Let $B$ be a basis of $A$. There exists a finite subset $C\subset B$ such that $A'\subset \bigoplus_{b\in C}\mathbb{Z}b$. We get an exact sequence:
\[ 0 \longrightarrow \Big(\bigoplus_{b\in C} \mathbb{Z}b\Big)\Big/A' \longrightarrow A/A' \longrightarrow \bigoplus_{b\in B\setminus C}\mathbb{Z}b \longrightarrow 0 \]
The left abelian group is finitely generated, particularly free-by-finite. The right group is free, hence also free-by-finite. By (ii), $A/A'\cong A''$ is free-by-finite as well.
\item It suffices to prove this assertion for finite groups and for $A=\mathbb{Z}$. For $A$ finite we have $nA = 0$ for $n = \#A$. Since $\mathbb{Z}$ has no nilpotent elements, the intersection of all prime ideals of $\mathbb{Z}$ is already trivial. \qedhere
\end{enumerate}
\end{proof}

@@

\begin{proof}
Let $U = X\setminus S$ and $P_1\in U$ be a closed point. We set:
\[ C_1 = \ker(r_{P_1}|_{K^S}: K^S \to k(P_1)^\times) \]
Since $r_{P_1}|_{k^\times}:k^\times \to k(P_1)^\times$ is injective, we have $C_1\cap k^\times = 1$ and the following map is injective:
\[ C_1 \to\prod_{P\in S}\mathbb{Z},\ x \mapsto (v_P(x))_{P\in S} \]
Therefore, $C_1$ is a free abelian group of finite rank. If $C_1 = 1$, then we set $T = \{P_1\}$ and we are done. Otherwise, let $x\in C_1,\ x\neq 1$. Then there exists an $a\in k^\times$ with the following properties:
\begin{enumerate}[(i)]
\item There exists a point $P_2\in X_0\setminus(S\cup\{P_1\})$ which lies above the prime of $k(x)$ corresponding to the polynomial $x-a\in k[x]$.
\item $a$ is not a root of unity.
\end{enumerate}
Clearly, (i) and (ii) hold for almost all $a\in k^\times$. We set $C_2 = \ker(r_{P_2}|_{C_1})\subset C_1$. $C_2$ is free abelian of finite rank as well. By requirement $r_{P_2}(x)=a$ and $a\in k^\times$ is not of finite order, thus $x\not\in C_2$ and $\operatorname{rank}(C_2) < \operatorname{rank}(C_1)$. By repeating this process until $C_r=1$, we get points $P_2,\ldots,P_r\in U_0$, such that the following map is injective:
\[ C_1\to\prod_{i=2}^{r} k(P_i)^\times,\ x\mapsto(r_{P_i}(x))_i \]
Finally for $T=\{P_1,\ldots,P_r\}\subset U_0$, the map $r_{S,T}$ is injective. For the second assertion, note the following isomorphism:
\[ \operatorname{coker}(r_{S,T}) \cong\operatorname{coker}\Big(K^S/k^\times \to \Big(\prod_{P\in T}k(P)^\times \Big)\Big/k^\times \Big) \]
Since $K^S/k^\times$ is generated by the uniformizers of the corresponding valuations in $S$, Lemma \ref{5.2} (iii) and Lemma \ref{5.3} (ii) show that the latter group is free-by-finite.
\end{proof}
